---
title: "CFE Tree Cover Study"
author: "Bradley Saul"
date: "`r Sys.Date()`"
output: html_document
bibliography: primary_study_analysis.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(stringr)
library(dplyr)
library(sp)
# Load source data ####

datadate <- "20190422"
ids      <- read.csv(sprintf("data/study_identifications_%s.csv", datadate),
                     stringsAsFactors = FALSE)
points   <- read.csv(sprintf("data/study_points_%s.csv", datadate), 
                     header = TRUE, stringsAsFactors = FALSE)
```

# Study Design

The design was based on the random point sampling described in @kaspar2017random. 2500 points each were selected in the rural buffer and county owned lands, plus 1500 points in the county outside the buffer and county lands. Using a [custom web application](https://github.com/bsaul/ocForestCover), users labeled each point as TREE, NOT TREE, or UNSURE based on NC Onemap [orthoimagery from 2008 and 2017](http://data.nconemap.gov/geoportal/catalog/raster/download.page). For power calculations, we assumed the study areas had 65% tree cover in 2008. Based on a type 1 error rate of 0.05, the study was designed to have 80% statistical power to detect a change in tree cover of ~3% within both the rural buffer and county lands. 

The figure below shows all the points selected in the study. The purple points are the rural buffer;  the red are county-owned lands; and the green are points in the county excluding the rural buffer or county-owned lands.

```{r, fig.height = 7}
spl_data <- readRDS(file = "data/study_spatial_data.rds")
primary_points_a <- readRDS(file = "data/primary_points.rds")
## Plotting ####
plot(spl_data[["oc"]])
plot(spl_data[["rb"]], add = TRUE)
plot(spl_data[["cl"]], add = TRUE)

filter(primary_points_a, area == "rb") %>%
  { points(.$lon,
           .$lat,
         pch = 3, cex = .1,
         col = "#bebada") }
filter(primary_points_a, area == "cl") %>%
  {points(.$lon,
         .$lat,
         pch = 3, cex = .1,
         col = "#fb8072")}
filter(primary_points_a, area == "oc") %>%
  {points(.$lon,
         .$lat,
         pch = 3, cex = .1,
         col = "#8dd3c7")}
```

The application measured intra- and interrater reliability by displaying points that had previously been identified by the same user or another user with a probablity of 0.15.

# Study Results

```{r}
# Create analytic dataset ####
analysis_ids <- ids %>%
  select(-X_rev) %>%
  filter(study_id == "oc_primary_study") %>%
  tidyr::separate(
    X_id, into = c("id", "point", "time"), sep = "_"
  ) %>%
  select(-db) %>%
  mutate(
    timestamp = lubridate::ymd_hms(timestamp)
  ) %>% 
  left_join(
    select(points, point = `X_id`, area),
    by = c("point")
  ) 

analysis_dt <- analysis_ids %>%
  group_by(point, area, time) %>%
  summarise(
    nids = n(),
    nT   = sum(value == "T"),
    nN   = sum(value == "N"),
    nU   = sum(value == "U")
  ) %>%
  mutate(
    agreement = (nids == nT) | (nids == nN) | (nids == nU),
    value     = case_when(
      agreement & (nT > 0) ~ "T",
      agreement & (nN > 0) ~ "N",
      agreement & (nU > 0) ~ "Ua",
      TRUE ~ "Ub"
    )
  ) 
```



```{r}
summary_by_point <- analysis_dt %>%
  group_by(area, point) %>%
  summarize(
    has_2008 = sum(time == "y2008") > 0,
    has_2017 = sum(time == "y2017") > 0,
    has_both = has_2017 & has_2008
  ) %>% group_by(area) %>%
  summarise(
    y2008 = sum(has_2008),
    y2017 = sum(has_2017),
    both  = sum(has_both)
  ) %>%
  {
    bind_rows(
      ., 
      summarise_if(., is.numeric, sum ) %>% mutate(area = "")
    ) 
  } %>%
  mutate(
    p = case_when(
      area == "cl" ~ both/2500,
      area == "rb" ~ both/2500,
      area == "oc" ~ both/1500,
      area == ""   ~ both/6500
    )
  )
```

`r length(unique(ids$uid))` people participated in the identification of points. In total, `r sum(analysis_dt$nids)` identifications were made, and `r summary_by_point$both[4]` points had at least one identification for both years (`r round(summary_by_point$p[4], 2) * 100`% of the available points).

```{r}
summary_by_point %>%
  mutate(
    area =case_when(
      area == "cl" ~ "Orange County owned property",
      area == "rb" ~ "Rural buffer",
      area == "oc" ~ "County excluding rural buffer and OC property",
      area == ""   ~ "Overall"
    )
  ) %>%
  knitr::kable(
    col.names = c("Area", "2008", "2017", "Both years", "Proportion of availabe points"),
    digits = 2
  )
```

## Rater Reliability

### Intrarater

```{r}
intra_counts <- analysis_ids %>%
  group_by(point, time, uid) %>%
  tally() %>% 
  group_by(n) %>%
  count() 

intra_counts %>%
  knitr::kable(
    col.names = c("# of times", "# of point/years"),
    caption   = "Number of times the same user identified the same point/year"
  )
```

In the `r intra_counts %>% filter(n > 1) %>% pull(nn) %>% sum()` point/years where a given user identified the same point/year more than once, the following table shows the number of point/years and the proportion of those point/years were each user agreed on all their identifications of that point/year.

```{r}
intra_by_user <- analysis_ids %>%
  group_by(point, time, uid) %>%
  filter(n() > 1) %>%
  group_by(point, time, uid) %>%
  arrange(value) %>%
  summarise(
    n         = n(),               
    agreement = all(value == value[1]),
    pattern   = paste0(value, collapse = "")
  ) 

intra_by_user %>%
  group_by(uid) %>%
  summarise(
    n         = n(),
    agreement = mean(agreement)
  ) %>%
  knitr::kable(
    col.names = c("User", "Number of point/years", "Proportion of Agreement"),
    digits    = 2,
    caption   = "Agreement per user"
  )

```


```{r}
intra_by_user %>%
  filter(!agreement) %>%
  group_by(pattern) %>%
  count() %>%
  knitr::kable(
    col.names = c("Pattern", "n"),
    caption   = "Patterns of Intrarater disagreement"
  )
```


### Interrater

```{r}
inter_count <- analysis_ids %>%
  group_by(point, time) %>%
  summarise(
    n_users = length(unique(uid))
  ) %>%
  group_by(n_users) %>%
  tally()

inter_count %>%
  knitr::kable(
    col.names = c("# of different users", "# of point/years"),
    caption   = "Number of users who identified a point/year"
  )
```

```{r}
inter_points <- analysis_ids %>%
  group_by(point, time, uid) %>%
  filter(n() == 1) %>%
  group_by(point, time) %>%
  filter(n() > 1) %>%
  arrange(value, .by_group = TRUE) %>%
  summarise(
    agreement = all(value == value[1]),
    pattern = paste0(value, collapse = "")
  )
```

Excluding the point/years considered for intrarater reliablity, the proportion of point/years where there was complete interrater reliability was `r inter_points$agreement %>% mean() %>% round(2)`.

```{r}
inter_points %>%
  filter(!agreement) %>%
  group_by(pattern) %>%
  tally() %>%
  knitr::kable(
    col.names = c("Patterns", "n"),
    caption   = "Patterns of Interrater Disagreement",
    digits    = 2
  )
```



## Paired Analysis

### Handling Disagreements

Based on the intra- and interrater reliablity assessments, which generally showed good agreement, the following algorithm was chosen for point/years with disagreement:

* the majority rules in cases where more than half of the identifications (by any user and/or multiple by the same user) were a particular value; e.g., "NNT" $\mapsto$ "N"; "TTU" $\mapsto$ "T"; "TUU" $\mapsto$ "U"; etc.
* all others are set to "U"

```{r}
vote <- function(x){
  
  if(length(x) == 1){
    return(x)
  } else {
    tab <- table(x)
    ptab <- prop.table(tab)
  
    if(any(ptab > 0.5)){
      names(tab)[which.max(ptab)]
    } else {
      return("U")
    }
  }
}

primary_analysis <- analysis_ids %>%
  select(point, area, time, value) %>%
  group_by(point, area, time) %>%
  summarise(
    value = vote(value)
  ) %>%
  group_by(point) %>%
  # Exclude points not IDed in both years
  filter(n() > 1) %>%
  ungroup()
```

### Results

The following table shows the observed pattern in `2008 -> 2017` tree cover in each of the three areas in our study after applying the voting algorithm described above.

```{r}
primary_analysis %>%
  tidyr::spread(key = time, value = value, fill = "U") %>%
  mutate(
    pattern = sprintf("%s -> %s", y2008, y2017)
  ) %>%
  group_by(area, pattern) %>%
  tally() %>%
  ungroup() %>%
  group_by(area) %>%
  mutate(
    p = n/sum(n)
  ) %>%
  select(area, pattern, p) %>%
  tidyr::spread(
    key = area, value = p
  ) %>%
  knitr::kable(
    col.names = c("Pattern", "OC property", "County Excluding RB and OC prop", "Rural Buffer"),
    digits    = 3,
    caption   = "Observed proportions of patterns of change in tree cover"
  )
```

Most often a "U" indicates that the user was unable to make an identification because an image failed to load rather than being uncertain about an identification. That is, a "U" identification is unlikely to depend on the actual state of a point or the user. Hence, we assume the "U" identifications are [missing completely at random](https://en.wikipedia.org/wiki/Missing_data#Missing_completely_at_random), and exclude all points with a "U" in either year is excluded from our primary analysis. 

The following table shows tree cover patterns after excluding these points.

```{r}
pad <- primary_analysis %>%
  tidyr::spread(key = time, value = value, fill = "U") %>%
  filter(y2008 != "U" & y2017 != "U") 

pad %>%
  mutate(
    pattern = sprintf("%s -> %s", y2008, y2017)
  ) %>%
  group_by(area, pattern) %>%
  tally() %>%
  ungroup() %>%
  group_by(area) %>%
  mutate(
    p = n/sum(n)
  ) %>%
  select(area, pattern, p) %>%
  tidyr::spread(
    key = area, value = p
  ) %>%
  knitr::kable(
    col.names = c("Pattern", "OC property", "County Excluding RB and OC prop", "Rural Buffer"),
    digits    = 3,
    caption   = "Observed Proportions of Patterns of change in tree cover"
  )


```

The following table shows the estimated change in tree cover for each study area with an adjusted Wald 95\% confidence interval for matched pairs [@agresti2005simple; @propci2018].


```{r}
pa_stats <- pad %>% 
  group_nest(area) %>%
  mutate(
    cont_tab   = purrr::map(data, ~ table(.x$y2017, .x$y2008)),
    diffpropci = purrr::map(
      .x = cont_tab, 
      .f = ~ PropCIs::diffpropci.mp(.x[1, 2], .x[2, 1], sum(.x), 0.05)),
    estimate   = purrr::map_dbl(diffpropci, ~ .x$estimate),
    conf_lo    = purrr::map_dbl(diffpropci, ~ .x$conf.int[1]),
    conf_hi    = purrr::map_dbl(diffpropci, ~ .x$conf.int[2])
  ) %>%
  select(area, estimate, conf_lo, conf_hi)

pa_stats %>%
  mutate(
    area = case_when(
      area == "cl" ~ "Orange County owned property",
      area == "rb" ~ "Rural buffer",
      area == "oc" ~ "County excluding rural buffer and OC property"),
    est = sprintf("%.3f (%.4f, %.4f)", estimate, conf_lo, conf_hi)
  ) %>%
  select(area, est) %>%
  knitr::kable(
    caption   = "Estimated change in tree cover 2008-2017",
    col.names = c("Area", "Estimate (95% CI)"),
    digits    = 3
  )
```



```{r, eval = FALSE}
primary_analysis %>%
  mutate(
    value = if_else(value == "U", "N", value)
  ) %>%
  tidyr::spread(key = time, value = value, fill = "U") %>%
  mutate(
    pattern = sprintf("%s -> %s", y2008, y2017)
  ) %>%
  group_by(area, pattern) %>%
  tally() %>%
  ungroup() %>%
  group_by(area) %>%
  mutate(
    p = n/sum(n)
  ) %>%
  select(area, pattern, p) %>%
  tidyr::spread(
    key = area, value = p
  ) %>%
  knitr::kable(
    col.names = c("Pattern", "OC property", "County Excluding RB and OC prop", "Rural Buffer"),
    digits    = 3,
    caption   = "Observed Proportions of Patterns of change in tree cover"
  )


```

# Comparison to other studies

@giorgino2018trends Table 3: 

```{r}
library(ggplot2)
sir20185077_table3 <- read.csv("data/sir20185077_table3.csv",
                               stringsAsFactors = FALSE)

sir20185077_table3 %>%
  filter(map_no != "") %>%
  select(map_no, in_orange, site_desc, drainage, delta_for, contains("pop")) %>%
  mutate(
    delta_pop_rel = (pop2010/pop1990 - 1) * 100
  ) %>%
  arrange(delta_for) %>%
  filter(
    grepl("S", map_no)
  ) -> plotdt

ggplot(
  data = plotdt %>% filter(delta_pop_rel < 250),
  aes(x = delta_pop_rel, y = -delta_for)
) + 
  geom_point(
    aes(color = factor(in_orange)),
    shape = 20
  ) +
  scale_color_manual(
    values = c("black", "orange"),
    guide = FALSE
  ) +
  scale_y_continuous(
    limits = c(0, 30),
    "% loss in forest land type (1991-2011)"
  ) + 
  scale_x_continuous(
    "% gain in population density (1990-2010)"
  ) + 
  labs(
    title   = "Trends in selected watersheds\nin the Triangle area of North Carolina",
    caption = "Watersheds contained within Orange County shown in orange.\nData are available in tables 2 and 3 of this USGS report: https://doi.org/10.3133/sir20185077.\nA watershed in the Cary area with 920% population density increase and 36% forest loss is not displayed."
  ) +
  theme_classic() + 
  theme(
    plot.caption = element_text(size = 6)
  )

```

# Summary

Our study shows a consistent pattern of increasing tree cover by 3-4% from 2008 to 2017 on Orange County owned property, the rural buffer surrounding Chapel Hill and Carrboro, and the rest of the county excluding the first two areas.



# Lessons Learned

* overlap probability too high
* exclude Lake Orange from county-owned property?
* send out regular updates on the status of the identifications
* app needs work to speed up image loading times

# References
