---
title: "CFE Tree Cover Study"
author: "Bradley Saul"
date: "`r Sys.Date()`"
output: html_document
bibliography: primary_study_analysis.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(stringr)
library(dplyr)
library(sp)
library(ggplot2)
# Load source data ####

datadate <- "20190422"
ids      <- read.csv(sprintf("data/study_identifications_%s.csv", datadate),
                     stringsAsFactors = FALSE)
points   <- read.csv(sprintf("data/study_points_%s.csv", datadate), 
                     header = TRUE, stringsAsFactors = FALSE)
```

# Study Design

The design was based on the random point sampling described in @kaspar2017random. 2500 points each were selected in the rural buffer and county owned lands, plus 1500 points in the county outside the buffer and county lands. Using a [custom web application](https://github.com/bsaul/ocForestCover), users labeled each point as TREE, NOT TREE, or UNSURE based on NC Onemap [orthoimagery from 2008 and 2017](http://data.nconemap.gov/geoportal/catalog/raster/download.page). For power calculations, we assumed the study areas had 65% tree cover in 2008. Based on a type 1 error rate of 0.05, the study was designed to have 80% statistical power to detect a change in tree cover of ~3% within both the rural buffer and county lands. 

The figure below shows all the points selected in the study. The purple points are the rural buffer;  the red are county-owned lands; and the green are points in the county excluding the rural buffer or county-owned lands.

```{r, fig.height = 7}
spl_data <- readRDS(file = "data/study_spatial_data.rds")
primary_points_a <- readRDS(file = "data/primary_points.rds")
## Plotting ####
plot(spl_data[["oc"]])
plot(spl_data[["rb"]], add = TRUE)
plot(spl_data[["cl"]], add = TRUE)

filter(primary_points_a, area == "rb") %>%
  { points(.$lon,
           .$lat,
         pch = 3, cex = .1,
         col = "#bebada") }
filter(primary_points_a, area == "cl") %>%
  {points(.$lon,
         .$lat,
         pch = 3, cex = .1,
         col = "#fb8072")}
filter(primary_points_a, area == "oc") %>%
  {points(.$lon,
         .$lat,
         pch = 3, cex = .1,
         col = "#8dd3c7")}
```

The application measured intra- and interrater reliability by displaying points that had previously been identified by the same user or another user with a probablity of 0.15.

# Study Results

```{r}
# Create analytic dataset ####
analysis_ids <- ids %>%
  select(-X_rev) %>%
  filter(study_id == "oc_primary_study") %>%
  tidyr::separate(
    X_id, into = c("id", "point", "time"), sep = "_"
  ) %>%
  select(-db) %>%
  mutate(
    timestamp = lubridate::ymd_hms(timestamp)
  ) %>% 
  left_join(
    select(points, point = `X_id`, area),
    by = c("point")
  ) 

analysis_dt <- analysis_ids %>%
  group_by(point, area, time) %>%
  summarise(
    nids = n(),
    nT   = sum(value == "T"),
    nN   = sum(value == "N"),
    nU   = sum(value == "U")
  ) %>%
  mutate(
    agreement = (nids == nT) | (nids == nN) | (nids == nU),
    value     = case_when(
      agreement & (nT > 0) ~ "T",
      agreement & (nN > 0) ~ "N",
      agreement & (nU > 0) ~ "Ua",
      TRUE ~ "Ub"
    )
  ) 
```



```{r}
summary_by_point <- analysis_dt %>%
  group_by(area, point) %>%
  summarize(
    has_2008 = sum(time == "y2008") > 0,
    has_2017 = sum(time == "y2017") > 0,
    has_both = has_2017 & has_2008
  ) %>% group_by(area) %>%
  summarise(
    y2008 = sum(has_2008),
    y2017 = sum(has_2017),
    both  = sum(has_both)
  ) %>%
  {
    bind_rows(
      ., 
      summarise_if(., is.numeric, sum ) %>% mutate(area = "")
    ) 
  } %>%
  mutate(
    p = case_when(
      area == "cl" ~ both/2500,
      area == "rb" ~ both/2500,
      area == "oc" ~ both/1500,
      area == ""   ~ both/6500
    )
  )
```

`r length(unique(ids$uid))` people participated in the identification of points. In total, `r sum(analysis_dt$nids)` identifications were made, and `r summary_by_point$both[4]` points had at least one identification for both years (`r round(summary_by_point$p[4], 2) * 100`% of the available points).

```{r}
summary_by_point %>%
  mutate(
    area =case_when(
      area == "cl" ~ "Orange County owned property",
      area == "rb" ~ "Rural buffer",
      area == "oc" ~ "County excluding rural buffer and OC property",
      area == ""   ~ "Overall"
    )
  ) %>%
  knitr::kable(
    col.names = c("Area", "2008", "2017", "Both years", "Proportion of availabe points"),
    digits = 2
  )
```

## Rater Reliability

### Intrarater

```{r}
intra_counts <- analysis_ids %>%
  group_by(point, time, uid) %>%
  tally() %>% 
  group_by(n) %>%
  count() 

intra_counts %>%
  knitr::kable(
    col.names = c("# of times", "# of point/years"),
    caption   = "Number of times the same user identified the same point/year"
  )
```

In the `r intra_counts %>% filter(n > 1) %>% pull(nn) %>% sum()` point/years where a given user identified the same point/year more than once, the following table shows the number of point/years and the proportion of those point/years were each user agreed on all their identifications of that point/year.

```{r}
intra_by_user <- analysis_ids %>%
  group_by(point, time, uid) %>%
  filter(n() > 1) %>%
  group_by(point, time, uid) %>%
  arrange(value) %>%
  summarise(
    n         = n(),               
    agreement = all(value == value[1]),
    pattern   = paste0(value, collapse = "")
  ) 

intra_by_user %>%
  group_by(uid) %>%
  summarise(
    n         = n(),
    agreement = mean(agreement)
  ) %>%
  knitr::kable(
    col.names = c("User", "Number of point/years", "Proportion of Agreement"),
    digits    = 2,
    caption   = "Agreement per user"
  )

```


```{r}
intra_by_user %>%
  filter(!agreement) %>%
  group_by(pattern) %>%
  count() %>%
  knitr::kable(
    col.names = c("Pattern", "n"),
    caption   = "Patterns of Intrarater disagreement"
  )
```


### Interrater

```{r}
inter_count <- analysis_ids %>%
  group_by(point, time) %>%
  summarise(
    n_users = length(unique(uid))
  ) %>%
  group_by(n_users) %>%
  tally()

inter_count %>%
  knitr::kable(
    col.names = c("# of different users", "# of point/years"),
    caption   = "Number of users who identified a point/year"
  )
```

```{r}
inter_points <- analysis_ids %>%
  group_by(point, time, uid) %>%
  filter(n() == 1) %>%
  group_by(point, time) %>%
  filter(n() > 1) %>%
  arrange(value, .by_group = TRUE) %>%
  summarise(
    agreement = all(value == value[1]),
    pattern = paste0(value, collapse = "")
  )
```

Excluding the point/years considered for intrarater reliablity, the proportion of point/years where there was complete interrater reliability was `r inter_points$agreement %>% mean() %>% round(2)`.

```{r}
inter_points %>%
  filter(!agreement) %>%
  group_by(pattern) %>%
  tally() %>%
  knitr::kable(
    col.names = c("Patterns", "n"),
    caption   = "Patterns of Interrater Disagreement",
    digits    = 2
  )
```



## Paired Analysis

### Handling Disagreements

Based on the intra- and interrater reliablity assessments, which generally showed good agreement, the following algorithm was chosen for point/years with disagreement:

* the majority rules in cases where more than half of the identifications (by any user and/or multiple by the same user) were a particular value; e.g., "NNT" $\mapsto$ "N"; "TTU" $\mapsto$ "T"; "TUU" $\mapsto$ "U"; etc.
* all others are set to "U"

```{r}
vote <- function(x){
  
  if(length(x) == 1){
    return(x)
  } else {
    tab <- table(x)
    ptab <- prop.table(tab)
  
    if(any(ptab > 0.5)){
      names(tab)[which.max(ptab)]
    } else {
      return("U")
    }
  }
}

primary_analysis <- analysis_ids %>%
  select(point, area, time, value) %>%
  group_by(point, area, time) %>%
  summarise(
    value = vote(value)
  ) %>%
  group_by(point) %>%
  # Exclude points not IDed in both years
  filter(n() > 1) %>%
  ungroup()
```

### Results

The following table shows the observed pattern in `2008 -> 2017` tree cover in each of the three areas in our study after applying the voting algorithm described above.

```{r}
primary_analysis %>%
  tidyr::spread(key = time, value = value, fill = "U") %>%
  mutate(
    pattern = sprintf("%s -> %s", y2008, y2017)
  ) %>%
  group_by(area, pattern) %>%
  tally() %>%
  ungroup() %>%
  group_by(area) %>%
  mutate(
    p = n/sum(n)
  ) %>%
  select(area, pattern, p) %>%
  tidyr::spread(
    key = area, value = p
  ) %>%
  knitr::kable(
    col.names = c("Pattern", "OC property", "County Excluding RB and OC prop", "Rural Buffer"),
    digits    = 3,
    caption   = "Observed proportions of patterns of change in tree cover"
  )
```

Most often a "U" indicates that the user was unable to make an identification because an image failed to load rather than being uncertain about an identification. That is, a "U" identification is unlikely to depend on the actual state of a point or the user. Hence, we assume the "U" identifications are [missing completely at random](https://en.wikipedia.org/wiki/Missing_data#Missing_completely_at_random), and exclude all points with a "U" in either year is excluded from our primary analysis. 

The following table shows tree cover patterns after excluding these points.

```{r primary_analysis_data}
pad <- primary_analysis %>%
  tidyr::spread(key = time, value = value, fill = "U") %>%
  filter(y2008 != "U" & y2017 != "U") 

pad %>%
  mutate(
    pattern = sprintf("%s -> %s", y2008, y2017)
  ) %>%
  group_by(area, pattern) %>%
  tally() %>%
  ungroup() %>%
  group_by(area) %>%
  mutate(
    p = n/sum(n)
  ) %>%
  select(area, pattern, p) %>%
  tidyr::spread(
    key = area, value = p
  ) %>%
  knitr::kable(
    col.names = c("Pattern", "OC property", "County Excluding RB and OC prop", "Rural Buffer"),
    digits    = 3,
    caption   = "Observed Proportions of Patterns of change in tree cover"
  )


```

The following table shows the estimated change in tree cover for each study area with an adjusted Wald 95\% confidence interval for matched pairs [@agresti2005simple; @propci2018].


```{r}
pa_stats <- pad %>% 
  group_nest(area) %>%
  mutate(
    cont_tab   = purrr::map(data, ~ table(.x$y2017, .x$y2008)),
    diffpropci = purrr::map(
      .x = cont_tab, 
      .f = ~ PropCIs::diffpropci.mp(.x[1, 2], .x[2, 1], sum(.x), 0.05)),
    estimate   = purrr::map_dbl(diffpropci, ~ .x$estimate),
    conf_lo    = purrr::map_dbl(diffpropci, ~ .x$conf.int[1]),
    conf_hi    = purrr::map_dbl(diffpropci, ~ .x$conf.int[2])
  ) %>%
  select(area, estimate, conf_lo, conf_hi)

pa_stats %>%
  mutate(
    area = case_when(
      area == "cl" ~ "Orange County owned property",
      area == "rb" ~ "Rural buffer",
      area == "oc" ~ "County excluding rural buffer and OC property"),
    est = sprintf("%.3f (%.4f, %.4f)", estimate, conf_lo, conf_hi)
  ) %>%
  select(area, est) %>%
  knitr::kable(
    caption   = "Estimated change in tree cover 2008-2017",
    col.names = c("Area", "Estimate (95% CI)"),
    digits    = 3
  )
```


```{r, eval = FALSE}
primary_analysis %>%
  mutate(
    value = if_else(value == "U", "N", value)
  ) %>%
  tidyr::spread(key = time, value = value, fill = "U") %>%
  mutate(
    pattern = sprintf("%s -> %s", y2008, y2017)
  ) %>%
  group_by(area, pattern) %>%
  tally() %>%
  ungroup() %>%
  group_by(area) %>%
  mutate(
    p = n/sum(n)
  ) %>%
  select(area, pattern, p) %>%
  tidyr::spread(
    key = area, value = p
  ) %>%
  knitr::kable(
    col.names = c("Pattern", "OC property", "County Excluding RB and OC prop", "Rural Buffer"),
    digits    = 3,
    caption   = "Observed Proportions of Patterns of change in tree cover"
  )


```

# Comparison to other studies

@giorgino2018trends Table 2 and 3 showed loss of forest land type in 3 OC watersheds from 1991-2011 by roughly 5-7%:

```{r}
library(ggplot2)
sir20185077_table3 <- read.csv("data/sir20185077_table3.csv",
                               stringsAsFactors = FALSE)

sir20185077_table3 %>%
  filter(map_no != "") %>%
  select(map_no, in_orange, site_desc, drainage, delta_for, contains("pop")) %>%
  mutate(
    delta_pop_rel = (pop2010/pop1990 - 1) * 100
  ) %>%
  arrange(delta_for) %>%
  filter(
    grepl("S", map_no)
  ) -> plotdt

orange_plot <- plotdt %>%
  dplyr::filter(in_orange == 1)
  

ggplot(
  data = plotdt %>% filter(delta_pop_rel < 250),
  aes(x = delta_pop_rel, y = delta_for)
) + 
  geom_hline(
    yintercept = 0
  ) + 
  geom_segment(
    data = orange_plot,
    aes(xend = 35, yend = -5),
    color = "grey50", size = 0.2
  ) + 
  geom_point(
    aes(color = factor(in_orange)),
    shape = 20
  ) +
  annotate("text", x = 35, y = -5, 
           label = "Orange County Watersheds",
           size = 2,
           vjust = -1) + 
  scale_color_manual(
    values = c("black", "orange"),
    guide = FALSE
  ) +
  scale_y_continuous(
    limits = c(-30, 0),
    expand = c(0, 0),
    "% loss in forest land type (1991-2011)"
  ) + 
  scale_x_continuous(
    "% gain in population density (1990-2010)"
  ) + 
  labs(
    title   = "Trends in selected stream watersheds\nin the Triangle area of North Carolina",
    caption = "Data are available in tables 2 and 3 of this USGS report: https://doi.org/10.3133/sir20185077.\nA watershed in the Cary area with 920% population density increase and 36% forest loss is not displayed."
  ) +
  theme_bw() + 
  theme(
    axis.line.x  = element_blank(), 
    plot.caption = element_text(size = 6)
  ) -> p

# ggsave("land_tree_cover_fig2.pdf", p, width = 6, height = 4)

p
```

# Validation 

```{r}
### Validation Study
set.seed(1234)
validation_points <- pad %>%
  group_by(area, y2008, y2017) %>%
  sample_frac(size = 0.05) %>%
  left_join(
    primary_points_a, by = c("point" = "_id", "area")
  ) 

# saveRDS(
#   validation_points,
#   file = sprintf("data/validation_study_%s.rds", format(Sys.Date(), "%Y%m%d")))

 
```

The following table shows the validation pattern (rows) vs the observed pattern (column):

```{r}
validation_res <- read.csv(
  file = "extdata/20190518_validation_responses.csv",
  stringsAsFactors = FALSE)

validation_res %>%
  select(point = Point, X2008, X2017) %>%
  inner_join(
    validation_points, by = c("point")
  ) %>%
  mutate(
    m2008 = (X2008 == y2008),
    m2017 = (X2017 == y2017),
    vpat = sprintf("%s->%s", X2008, X2017),
    opat = sprintf("%s->%s", y2008, y2017),
    match = (vpat  == opat)
  )  %>%
  filter(vpat != "N->U") %>%
  group_by(vpat, opat) ->
  hold
vmargin <- table(hold$vpat) %>% prop.table()
# hold %>%
#   group_by(area, vpat) %>%
#   summarise(
#     m2008 = mean(m2008),
#     m2017 = mean(m2017)
#   )
# table(v = hold$vpat) %>% prop.table()
# table(o = hold$opat) %>% prop.table()
table(v = hold$vpat, o = hold$opat) -> x
x

```

Similar to our interrater analysis, there was complete agreement in `r round(sum(diag(x))/sum(x), 2)` of the validation sample. The change in tree cover (across all areas) estimated by the validation sample was `r round(vmargin[2] - vmargin[3], 3)` The most common disagreement was original = `N->T` and validation = `T->T`, which does suggest a bias toward 2008 images being incorrectly identified as `N` when in fact they were `T`. This measurement error could shift our change in tree cover estimates toward 0. Given that we see that other types of measurement errors also occurred, in all, the pattern of increasing tree cover appears to hold, though the true rate may be less than estimated above due to measurement error.

Futher analyses could use (e.g.) Hidden Markov Models to try to account for measurement error treating the underlying state of a point as a latent state or an approach such as @magder1997logistic could be considered. This is beyond the scope of this analysis.

# Summary

Our study suggests increasing tree cover from 2008 to 2017 on Orange County owned property, the rural buffer surrounding Chapel Hill and Carrboro, and the rest of the county excluding the first two areas.

# Lessons Learned

* overlap probability unnecessarily high
* exclude Lake Orange from county-owned property?
* send out regular updates on the status of the identifications
* app needs work to speed up image loading times

# References
